Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-01-21 analysis/paper/article.qmd
JB edits (final, hopefully, excepting references) on 21 January 2025

Success in tenure-track job applications in archaeology is strongly determined by where applicants get their PhDs. Data from the 2014–2015 AnthroGuide publication of the American Anthropological Association shows that just ten out of over 100 US graduate programs produced over 30% of the graduates hired into tenure-track faculty positions [@speakman2018choosing]. Top-ranking programs are placing fewer than one in three graduates in tenure-track jobs [@mackieMarketShareAccounting2023]. Similarly, a network analysis of 1,918 faculty holding tenured or tenure-track positions at PhD-granting anthropology programs in the US in 2015 (including 506 archaeologists) found that the just fifteen graduate programs produced 52.8 percent of tenured and tenure-track positions [@kawaSocialNetworkUS2019]. This network analysis showed that programs with large endowments, with faculty who hold prestigious awards and are widely cited by other scholars, produce the majority of tenured and tenure-track faculty in anthropology. Hiring bias predicated on the prestige of specific institutions and programs is not unique to anthropology. Targeted studies of the disciplines of sociology, communication, and operations research, and industrial systems engineering show similar trends (**Barnett et al., 2010; Castillo et al., 2018; Feeley and Tutzauer 2022; Nevin 2019**). **Broader analyses comparing X, Y and Z, and analyses while a recent analysis of XXX showed Z (Clauset et al., 2015;** [@wapmanQuantifyingHierarchyDynamics2022a]). Within anthropology, however, such studies draw attention to the paradox of a discipline committed to fighting social inequalities continuing to reproduce its own systematic inequalities through hiring practices that favor an elite minority of graduates with prestigious institutional affiliations.

@@

**While attention to equity in job market materials is commendable, work on hiring practices suggests ZZZ. For example, Warren 2019 has shown XXX. Rivera, in a study of hiring practices at YYY, has likewise demonstrated ZZZ. Somewhere in this section I want to discuss the results from sociology, specifically the Warren 2019 piece and the Rivera 2017 piece on hiring practices.** <!-- ok, sounds good! -->

**Cite Rogge 1976.**

@@

Our study of a decade of tenure-track job ads in archaeology in the US reveals diverse dynamics in the demand for specialisations in topics, methods and geographic focus, and in the instructions to applicants to these jobs <!--What shifts? It seems like there is an introductory sentence missing here. BM: ok, I tried to expand it a bit. JB: Yes, this flows now. Thx. -->. While these dynamics are familiar to scholars actively seeking faculty jobs, we believe this is the first time they have been quantified at such a large scale within archaeology. Trends in job ads reflect broader shifts in intellectual and practical priorities concerning archaeology, undergraduate education, and the process of hiring professors. The demand for archaeology faculty, indicated by the total number of tenure-track jobs, maybe explained by a variety of processes. Overall, we found more tenure-track jobs advertised each year than non-tenure-track, with the exception of 2013-14. This is at odds with previous work that has shown that contingent positions make up more than half of faculty positions [@colby2022annual]. **MORE DETAIL HERE FROM JB.** This discrepancy may be due to the more limited circulation of advertising for short-term positions relative to advertising for tenure-track jobs. Many of these short-term positions are not advertised nationally, but only through local email lists and are filled by people close to the hiring department, such as recently graduated students. One possible explanation for our results is that non-tenure-track jobs are less frequently added to the Academic Jobs Wiki. The downward trend in tenure-track positions during 2013-2019 may be related to declining undergraduate enrollment in anthropology fields since 2013 [@crambChangingProfileTenureTrack2022a]. The big dip during 2020-2021 is explained by the hiring freezes at many institutions resulting from the COVID-19 pandemic, which caused extreme disruption and uncertainty as universities focused on adapting to online instruction and assessment in an effort to minimize the spread of the virus [@woolstonJuniorResearchersHit2020]. A survey of early career researchers in archaeology captures the impact of this dip, with three-quarters of respondents experiencing negative impacts on their careers due to the pandemic [@bramiPrecariousFutureReflections2023]

@@

Our data suggest that archaeology faculty at US universities participated in this movement by adjusting their hiring plans to prioritize recruiting archaeologists working on topics relevant to Black and Indigenous communities. Many universities may have hoped to hire Black and Indigenous archaeologists as part of their effort to tackle systemic racism. However as it is illegal in the US for universities use race or ethnicity as a primary factor in hiring (due to the 1964 Civil Rights Act which prohibits hiring based solely on race or ethnicity), universities appear to have tailored the content of their job ads to focus on topics where they expect Black and Indigenous researchers to be most numerous, in an effort to ensure that these researchers would be well-represented in the pool of applicants. <!-- Some commentary on whether and if Black/Indigenous researchers are more numerous in these fields, citing Maria Franklin, Sonya Atalay, Aja, Rachel Watson, etc. BM: I've looked in these articles and not found any relevant data. Also, I think this is going beyond the scope of our study -->

This striking change in topics and geographic foci of 2019-2021 occurred against a backdrop of several longer-term trends. Over the course of the study period, we document a gradual decline in topics such as the archaeology of complex societies and archaeological science, the goegraphic foci of Mesoamerica and South America, and methods relating to landscape archaeology. These trends are harder to explain as we cannot link their origins to a moment like COVID-19. We might speculate that a growing preference for archaeological approaches that privilege agency-driven, relational perspectives, in combination with human subjectivity and cultural framing, is one enduring legacy of debates in the 1980s and 90s about processualism versus post-processualism [@hodderArchaeologicalProcessIntroduction1999; @johnson2019archaeological] <!-- JB: (Citations) BM: I added some, not my strong area, however -->. This might explain why archaeological science is showing a decline, as demand for methods for analyzing artefact materiality, ontology, and power displace physical laboratory methods for technological, functional, and compositional analyses. Other factors relevant to this decline may include increasing difficulty of obtaining research funding to support archaeological science research, such as laboratory facilities and instrumentation. A decline in interest in the archaeology of complex societies may reflect several themes that intersect with broader social changes, such as growing interest in Indigenous and non-state actors in the past, increased interest in climate change, environmental sustainability, and resilience, which may have shifted attention away from the study of monumental architecture, elite societies, political hierarchies and state systems.

@@

Our data on the requirements for applicants support prior findings that the complexity of applications—and concomitantly, the labor required to apply for a tenure-track job—has gradually increased over time. This trend is especially pronounced for Assistant Professor positions, which make more demands on applicants than Associate and Full Professor positions. Consistent with results from other studies \[\@HelloTristesTropes2021\], our study documented a growing demand for research and teaching statements. Demand for diversity statements shows a unique trajectory, peaking in 2020-2021 and declining into the present. This may relate to the intersecting concerns emerging during the COVID-19 pandemic, which may have reached a peak in 2020-2021 and then declined over time, resulting in dropping requests for diversity statements, as the most urgent period of the pandemic moves into the past. Another factor here may be the debates surrounding the experiment with diversity statements in hiring at some UC campuses during 2016-2022. Universities that previously required a diversity statement dropped that requirement after 2022 [@guidenAreDiversityStatements]. In our data we observed a reversal of the diversity statement requirement at `r n_schools_that_dropped_ds_req` universities. Those schools posted job ads prior 2022 that did require a diversity statement and also posted an ad in 2022 that did not require one.

@@

We thank NO ONE. <!-- haha! so generous of you! JB: We should probably add some people here, no? -->

>>>>

Success in tenure-track job applications in archaeology is strongly determined by where applicants get their PhDs. Data from the 2014–2015 AnthroGuide publication of the American Anthropological Association shows that just ten out of over 100 US graduate programs produced over 30% of the graduates hired into tenure-track faculty positions [@speakman2018choosing]. Top-ranking programs are placing fewer than one in three graduates in tenure-track jobs [@mackieMarketShareAccounting2023]. Similarly, a network analysis of 1,918 faculty holding tenured or tenure-track positions at PhD-granting anthropology programs in the US in 2015 (including 506 archaeologists) found that the just fifteen graduate programs produced 52.8 percent of tenured and tenure-track positions [@kawaSocialNetworkUS2019]. This network analysis showed that programs with large endowments, with faculty who hold prestigious awards and are widely cited by other scholars, produce the majority of tenured and tenure-track faculty. Hiring bias predicated on the prestige of specific institutions and programs is not unique to anthropology. Targeted studies of the disciplines of sociology, communication, and operations research, and industrial systems engineering show similar dynamics (**Barnett et al., 2010; Castillo, Myers and Chen, 2018; Feeley and Tutzauer 2022; Nevin 2019**). Broader analyses comparing computer science, business, and history likewise reveal \[parallel\] patterns among disparate disciplines across the humanities, social sciences, and STEM fields, in which "across disciplines, prestige hierarchies make the most accurate predictions of faculty placement" (Clauset et al., 2015:4). The most recent comprehensive meta-analysis of faculty placement dynamics in the United States, which examined employment and doctoral education of all tenure-track faculty at PhD-granting universities, underscored the outsized impact of institutional prestige: 80% of faculty trained in the US came from just 20.4% of institutions [@wapmanQuantifyingHierarchyDynamics2022a]. Within anthropology, however, such studies draw attention to the paradox of a discipline committed to fighting social inequalities continuing to reproduce its own systematic inequalities through hiring practices that favor an elite minority of graduates with prestigious institutional affiliations.

@@

@@

Our study of a decade of tenure-track job ads in archaeology in the US reveals diverse dynamics in the demand for specialisations in topics, methods and geographic focus, and in the instructions to applicants to these jobs <!--What shifts? It seems like there is an introductory sentence missing here. BM: ok, I tried to expand it a bit. JB: Yes, this flows now. Thx. -->. While these dynamics are familiar to scholars actively seeking faculty jobs, we believe this is the first time they have been quantified at such a large scale within archaeology. Trends in job ads reflect broader shifts in intellectual and practical priorities concerning archaeology, undergraduate education, and the process of hiring professors. The demand for archaeology faculty, indicated by the total number of tenure-track jobs, maybe explained by a variety of processes. Overall, we found more tenure-track jobs advertised each year than non-tenure-track, with the exception of 2013-14. This is at odds with previous work that has shown that contingent positions make up more than half of faculty positions [@colby2022annual], and that these numbers are increasing between 2004 and 2019, there was a 36% increase in the replacement of tenure-track positions by contingent positions (**AAUP 2022**). This discrepancy in our data set may be due to the more limited circulation of advertising for short-term positions relative to advertising for tenure-track jobs. Many of these short-term positions are not advertised nationally, but only through local email lists and are filled by people close to the hiring department, such as recently graduated students. One possible explanation for our results is that non-tenure-track jobs are less frequently added to the Academic Jobs Wiki. The downward trend in tenure-track positions during 2013-2019 may be related to declining undergraduate enrollment in anthropology fields since 2013 [@crambChangingProfileTenureTrack2022a]. The big dip during 2020-2021 is explained by the hiring freezes at many institutions resulting from the COVID-19 pandemic, which caused extreme disruption and uncertainty as universities focused on adapting to online instruction and assessment in an effort to minimize the spread of the virus [@woolstonJuniorResearchersHit2020]. A survey of early career researchers in archaeology captures the impact of this dip, with three-quarters of respondents experiencing negative impacts on their careers due to the pandemic [@bramiPrecariousFutureReflections2023]

@@

Our data suggest that archaeology faculty at US universities participated in this movement by adjusting their hiring plans to prioritize recruiting archaeologists working on topics relevant to Black and Indigenous communities. Many universities may have hoped to hire Black and Indigenous archaeologists as part of their effort to tackle systemic racism. However as it is illegal in the US for universities use race or ethnicity as a primary factor in hiring (due to the 1964 Civil Rights Act which prohibits hiring based solely on race or ethnicity), universities appear to have tailored the content of their job ads to focus on topics where they expect Black and Indigenous researchers to be most numerous, in an effort to ensure that these researchers would be well-represented in the pool of applicants. This striking change in topics and geographic foci of 2019-2021 occurred against a backdrop of several longer-term trends. Over the course of the study period, we document a gradual decline in topics such as the archaeology of complex societies and archaeological science, the goegraphic foci of Mesoamerica and South America, and methods relating to landscape archaeology. These trends are harder to explain as we cannot link their origins to a moment like COVID-19. We might speculate that a growing preference for archaeological approaches that privilege agency-driven, relational perspectives, in combination with human subjectivity and cultural framing, is one enduring legacy of debates in the 1980s and 90s about processualism versus post-processualism [@hodderArchaeologicalProcessIntroduction1999; @johnson2019archaeological] <!-- JB: (Citations) BM: I added some, not my strong area, however. JB: Do you think we could just cite Lars Fogelin's arch theory book here?-->. This might explain why archaeological science is showing a decline, as demand for methods for analyzing artefact materiality, ontology, and power displace physical laboratory methods for technological, functional, and compositional analyses. Other factors relevant to this decline may include increasing difficulty of obtaining research funding to support archaeological science research, such as laboratory facilities and instrumentation. A decline in interest in the archaeology of complex societies may reflect several themes that intersect with broader social changes, such as growing interest in Indigenous and non-state actors in the past, increased interest in climate change, environmental sustainability, and resilience, which may have shifted attention away from the study of monumental architecture, elite societies, political hierarchies and state systems.

@@

Our data on the requirements for applicants support prior findings that the complexity of applications—and concomitantly, the labor required to apply for a tenure-track job—has gradually increased over time. This trend is especially pronounced for Assistant Professor positions, which make more demands on applicants than Associate and Full Professor positions. Consistent with results from other studies \[\@HelloTristesTropes2021\] <!-- JB: Ben, this does not seem to be "hyperlinked" the way other references are, just flagging. -->, our study documented a growing demand for research and teaching statements. Demand for diversity statements shows a unique trajectory, peaking in 2020-2021 and declining into the present. This may relate to the intersecting concerns emerging during the COVID-19 pandemic, which may have reached a peak in 2020-2021 and then declined over time, resulting in dropping requests for diversity statements, as the most urgent period of the pandemic moves into the past. Another factor here may be the debates surrounding the experiment with diversity statements in hiring at some UC campuses during 2016-2022. Universities that previously required a diversity statement dropped that requirement after 2022 [@guidenAreDiversityStatements]. In our data we observed a reversal of the diversity statement requirement at `r n_schools_that_dropped_ds_req` universities. Those schools posted job ads prior 2022 that did require a diversity statement and also posted an ad in 2022 that did not require one.

@@

<!--JB: We should probably add some people here, no? -->
